---
title: "Iris Data Cleaning"
output: html_notebook
---

## Install the required packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install.packages('tidyverse')
# install.packages('gridExtra')
# install.packages('forcats')
#install.packages('viridis')
#install.packages('class')
#install.packages('modelr')
#install.packages('kknn')
#install.packages('pbkrtest')
#install.packages('caret')

```

## Import the required packages 
```{r}
library(tidyverse)
library(grid)
library(gridExtra)
library(forcats)
library(modelr)
library(caret)
library(kknn)
```

# Data Cleaning
1) Load the iris data into a tibble
2) Get an overview of the data
3) Coerce characters (strings) into factors (categorical data representation)
4) Reformat tibble into a LONG format (for ease of visualisation)

## Load Data into Tibble & Overview
```{r}
iris <- as_tibble(iris)
summary(iris)
```

## Explore the number of variables per class visually
```{r}
ggplot(iris, mapping = aes())+
  geom_bar(mapping = aes(x=Species,fill=Species))
```

## sapply() to check the class of variables
```{r}

sapply(iris,class)

```

ALSO could use the map() function (from the tidyverse)

```{r}
map(iris,sd)
map(iris,mean)
```

## how many na values?
```{r}
# CREATE A FUNCTION
na_data <- function(x){
  sum(is.na(x))/length(x)*100
}
#WHY DOES IT NEED THE /LENGTH*100??

na_data2 <- function(x){
  sum(is.na(x))
}

#loop it over our dataset
apply(long_iris,2,na_data2)

```

## Reformat the tibble into LONG format
```{r}
long_iris <- iris %>% 
  gather(key= 'part', value = 'value', Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) %>%
  separate(part,c('part','measure'), sep = '\\.')
```

gather() - key = the column header for the columns (as row values) / value = the name of the values (the old rows) / the columns you want under the key 
separate() - split the 'part' column at the '.' into two columns - 'part' and 'measure'
note: needs to be \\. to escape the first \ and then \.

## Coerce Characters into factors
```{r}
factors <- c('part','measure')
long_iris[factors] <- lapply(long_iris[factors],as.factor)
```
