---
title: "Data Exploration"
output: html_notebook
---

## Useful Function
Downloaded a useful function for viewing multiple plots
http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/ 

```{r}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

# Univariate Plots 

## Boxplot of Sepal Width Separated by Species

```{r}
ggplot(iris, mapping = aes(fill=Species)) +
  geom_boxplot(mapping = aes(x=iris$Species,y=iris$Sepal.Width))
```

## Frequency Polygons of each Variable

```{r}
P1 <- ggplot(iris, mapping = aes(x = Sepal.Length, color = Species, fill=Species)) +
  geom_freqpoly(binwidth=0.3) +
  theme(legend.position = 'none')

P2 <- ggplot(iris, mapping = aes(x = Sepal.Width, color = Species, fill=Species)) +
  geom_freqpoly(binwidth=0.3) +
  theme(legend.position = 'none')

P3 <- ggplot(iris, mapping = aes(x = Petal.Length, color = Species, fill=Species)) +
  geom_freqpoly(binwidth=0.3) +
  theme(legend.position = 'none')

P4 <- ggplot(iris, mapping = aes(x = Petal.Width, color = Species, fill=Species)) +
  geom_freqpoly(binwidth=0.3)

multiplot(P1, P2, P3, P4, cols=2)
```

## Density Plots of each Variable (using grid.arrange())

```{r}

P1 <- ggplot(iris, mapping = aes(x = Sepal.Length, color = Species, fill=Species)) +
  geom_density(alpha=0.3) +
  theme(legend.position = 'none')

P2 <- ggplot(iris, mapping = aes(x = Petal.Length, color = Species, fill=Species)) +
  geom_density(alpha=0.3) +
  theme(legend.position = 'none')

P3 <- ggplot(iris, mapping = aes(x = Sepal.Width, color = Species, fill=Species)) +
  geom_density(alpha=0.3) +
  theme(legend.position = 'none')

P4 <- ggplot(iris, mapping = aes(x = Petal.Width, color = Species, fill=Species)) +
  geom_density(alpha=0.3) +
  theme(legend.position = 'none')

grid.arrange(P1,
             P2,
             P3,
             P4,
             nrow = 2,
             top = textGrob("Iris Density Plot", gp=gpar(fontsize=10)))

```

## Use the LONG TABLE to show multiple plots of each Feature/Variable

```{r}
ggplot(data= long_iris, mapping= aes(x = Species, y = value, col = Species)) +
  geom_jitter(alpha=0.3, size=0.8) + 
  stat_boxplot(alpha = 0.5) + 
  facet_grid(part~measure) +
  ggtitle('Iris: Feature Exploration')
```

## Mean and Variance Exploration for each Feature

### First CREATE the Standard Deviation and Ymin/Ymax values for the error bars
```{r}
iris_mean_se <- long_iris %>%
  group_by(Species,part,measure) %>%
  summarise(
    mean = mean(value),
    sd = sd(value),
    ymax = mean(value) + sd(value),
    ymin = mean(value) - sd(value)
  )
iris_mean_se
```

### Then PLOT these Values ontop of MEAN BARS 
- NOTE you need to set the positition dodge to the same to get them to line up (0.9 = default value for bars)
- also necessary to place fill=Species in the entire ggplot mapping, otherwise only in local environment and not passed to geom_errorbar
```{r}

#USE MULTIPLOT to have multiple plots of the different subsets of the data 
dodge <- position_dodge(width = 0.9)

petal <- ggplot(data = subset(iris_mean_se, part=='Petal'), mapping = aes(x=measure,y=mean,fill = Species)) +
  geom_bar(stat='identity', position= dodge) + 
  geom_errorbar(mapping = aes(ymin= ymin, ymax= ymax), position= dodge, width = 0.2) +
  theme(legend.position = 'none')+
  ggtitle("Iris: Petal")

sepal <- ggplot(data = subset(iris_mean_se, part=='Sepal'), mapping = aes(x=measure,y=mean,fill = Species)) +
  geom_bar(stat='identity', position= dodge) + 
  geom_errorbar(mapping = aes(ymin= ymin, ymax= ymax), position= dodge, width = 0.2) +
  ggtitle("Iris: Sepal")

multiplot(petal, sepal, cols=2)

```

# Multi Variate Plots

## Scatter plot of Petals & Sepals (separated by Species)

```{r}
sepal_plot <- ggplot(iris) +
  geom_point(mapping=aes(x=Sepal.Length,y=Sepal.Width,col=Species,shape=Species)) +
  theme(legend.position = 'none')

petal_plot <- ggplot(iris) +
  geom_point(mapping=aes(x=Petal.Length,y=Petal.Width,col=Species,shape=Species)) +
  theme(legend.position = 'top')

multiplot(sepal_plot, petal_plot, cols=2)

```