---
title: "Model the Iris Data (Linear Regression)"
output: html_notebook
---
# 1) Linear Regression Model

## Create a WIDE TABLE Format 
FOR EACH Observation
1) Add flower # as a variable
2) have the PART of the flower as a categorical variable
3) get the LENGTH and WIDTH as the two variable with numerical values in them

TIDY Data 
  Each column a VARIABLE; 
  Each row an OBSERVATION;
  Each cell a VALUE;

```{r}
wide_iris <- iris %>%
  gather(key='key', value='value', -Species, -Flower) %>%
  separate(key, c('Part','Measure'), sep= "\\.") %>%
  spread(Measure, value)

wide_iris
```

### Explanation
http://r4ds.had.co.nz/tidy-data.html#spreading-and-gathering 
gather()
- select a key which is the NAME OF THE COLUMNS in the untidy data (current variables)
- the value for that variable 
-keep only SPECIES, FLOWER columns and then the variable being measured (key=) and then value for that observation (value=)
separate()
-split a column into two 
-split the key column, into 'part' and 'measure' at the '.' sign
spread()
-then spread the measure column into its separate components (length and width)
-taking the value column as the valu for each of those cells!
WE have tidy data :) 

## Build the Linear Model (using ggplot)
NOTE: This analysis was flawed by the fact that I didn't separate Petal or Sepal when comparing length and width in plot1 & plot3

### Load the ggplot() header
```{r}
lm_iris <- ggplot(data = wide_iris, mapping= aes(x= Width, y= Length, col=Species))
```

### Compare LENGTH and WIDTH for each species
```{r}
lm_iris1 <- lm_iris+
  geom_jitter(alpha= 0.4, size= 0.8) +
  facet_grid(. ~ Species) +
  stat_smooth(method= 'lm', se=F) + 
  ggtitle("Iris: Length vs. Width by Species")

lm_iris1
```

### Compare Petal Length vs. Petal Width & Sepal Width vs. Sepal Length 

```{r}
lm_iris2 <- ggplot(data=wide_iris, mapping=aes(x=Width, y=Length)) +
  geom_jitter(alpha=0.4, size=0.8) +
  facet_grid(.~ Part) +
  stat_smooth(method= 'lm', se=T) + 
  ggtitle("Iris: Length vs. Width by Flower Part")
  
lm_iris2
```

### How does length and width compare for each species 
NOTE: False Analysis because aggregated Sepals and Petals
```{r}
lm_iris3 <- lm_iris +
  geom_point(alpha = 0.4, size= 0.8) + stat_smooth(method = 'lm', fullrange= T, size= 0.5) +
  ggtitle("Iris: Length vs. Width by Species 2")
  
lm_iris3
```
